\section{Conclusion}\label{sec:conclusion}

This study investigated whether a single spatio-temporal framework can jointly forecast next-day regional evening-peak demand and a graph-level Operational Stress Index from Bangladesh national grid records under chronological evaluation. The Correlation-Aware Multi-Task Forecasting Framework---implemented as correlation-graph PF-STGT with parallel spatial and temporal transformer encoding and repaired multi-task loss---delivers the strongest joint credentials in the frozen benchmark suite: macro demand MAE 88.65~MW and OSI $R^2$ 0.745, with statistically supported demand improvement over the historical hybrid reference.

Five outcomes define the contribution. First, one checkpoint supplies coupled day-ahead load and stress forecasts from shared spatio-temporal representations. Second, correlation adjacency outperforms geographical priors on this dataset. Third, a documented demand--stress trade-off exists against demand-only training. Fourth, regional error geography is predictable, with meaningful Dhaka gains despite persistent concentration at the national load centre. Fifth, multi-method attribution provides coalition-level transparency with explicit discordance reporting.

The work extends spatio-temporal forecasting to coupled stress indices on a previously unreported Bangladesh division-level case \cite{Hossain2025BangladeshSmartGrid}, demonstrating that graph priors and task weighting require empirical tuning rather than domain intuition. Coupled day-ahead load-and-stress forecasting is achievable as a single-model service when graph structure, multi-task calibration, and division-level monitoring are co-designed with the evaluation evidence---appropriately qualified as offline evaluation rather than operational deployment.
