\section{Methodology}\label{sec:methodology}

\subsection{Problem Definition and Task Formulation}\label{subsection:problem_definition_and_task_formulation}

Operators of modern power systems must balance generation dispatch, reserve allocation, and network security under growing load and supply variability. Historical operational data motivate spatio-temporal frameworks that exploit regional load coupling and system-wide context. Operators also need visibility into operational stress from reserve shortfall, fuel and infrastructure limitations, and load-shedding pressure. This study addresses both requirements on the Bangladesh national power system; dataset properties appear in Section~\ref{subsection:dataset_description}.

We formulate day-ahead forecasting as a supervised multi-task learning problem \cite{Caruana1997Multitask} over a set of spatial units indexed by graph nodes. Let \(\mathcal{V} = \{1, \ldots, N\}\) denote the node set, with \(N\) regional divisions. Time is indexed in daily steps \(t\). At each forecast origin \(t\), the model receives a lookback window of length \(T\) consecutive observed days and predicts targets at horizon \(H\). Inputs and outputs are defined over the window \([t - T + 1, t]\) for observations and at \(t + H\) for supervision. The batch dimension is omitted where clear from context.

Inputs comprise two coupled tensors. Regional node features are denoted \(\mathbf{X}^{\mathrm{node}} \in \mathbb{R}^{T \times N \times F_n}\), where \(F_n\) is the number of node-level features per region per day, encoding division-level historical demand signals, engineered lag and rolling statistics, and related regional observables. National and system-level context is represented by \(\mathbf{X}^{\mathrm{global}} \in \mathbb{R}^{T \times F_g}\), where \(F_g\) is the number of global features per day, including calendar indicators, grid aggregates, operational limitation variables, and national generation scalars. Together, \(\mathbf{X}^{\mathrm{node}}\) and \(\mathbf{X}^{\mathrm{global}}\) represent all information available at or before day \(t\) from which forecasts are produced.

Task 1 produces a regional demand vector \(\hat{\mathbf{y}}^{\mathrm{demand}} = (\hat{D}_1, \ldots, \hat{D}_N)^\top \in \mathbb{R}^N\), with each \(\hat{D}_r\) expressed in megawatts (MW). Task 2 produces a graph-level operational stress scalar \(\hat{y}^{\mathrm{osi}} \in [0, 1]\). The demand output assigns one next-day prediction per node; the stress output is a single system-wide score rather than a per-region vector.

For Task 1, each node \(r \in \mathcal{V}\) predicts next-day regional evening-peak demand \(D_r(t+H)\) in MW. Node-level targets capture distinct load scales and spatially coupled dynamics that a national aggregate cannot represent. Regional supply is excluded because prior analysis documented strong demand--supply collinearity; division-level shedding is not regressed directly because it is sparse and zero-inflated at the regional level. Shedding intensity enters instead through the composite stress target below.

Task 2 predicts the Operational Stress Index (OSI), a bounded system-health score integrating shedding severity, generation reserve margin, and operational limitation pressure. OSI combines three components normalised using train-split bounds only:

\begin{align}
c_1 &= \frac{L_{\mathrm{total}}}{D_{\mathrm{total}}}, &
c_2 &= 1 - \frac{GR}{\mathrm{Highest\_Generation}}, \nonumber\\
c_3 &= \frac{\mathrm{TOL}}{\mathrm{Highest\_Generation}},
\end{align}

\begin{multline}
\mathrm{OSI} = \mathrm{mean}\bigl(\mathrm{minmax}_{\mathrm{train}}(c_1),\, \mathrm{minmax}_{\mathrm{train}}(c_2),\\
\mathrm{minmax}_{\mathrm{train}}(c_3)\bigr),
\end{multline}

where \(L_{\mathrm{total}}\) is total regional shedding (MW), \(D_{\mathrm{total}}\) is total regional demand, \(GR\) is generation reserve, and \(\mathrm{TOL}\) is total operational limitation. Continuous OSI regression preserves graded stress magnitude, including reserve- and limitation-driven pressure without a shedding event. The supervised target is \(\mathrm{OSI}(t+H)\).

Regional demand and OSI are learned jointly because the targets are operationally related yet statistically non-redundant. Demand forecasts describe expected load levels across spatial units, whereas OSI summarises constraint-related system pressure through components \(c_1\)--\(c_3\) that are not fully encoded by demand alone. Joint estimation provides synchronised day-ahead visibility over load and stress from shared calendar and system context. A leakage policy applies to stress supervision: when predicting \(\mathrm{OSI}(t+H)\), same-day \(\mathrm{OSI}(t)\) is excluded from the input feature set; only information observable at or before \(t\) may be used.

Both tasks use a single-step horizon \(H = 1\) day. At origin \(t\), the model predicts \(\{D_r(t+1)\}_{r \in \mathcal{V}}\) and \(\mathrm{OSI}(t+1)\) from inputs through day \(t\), aligning with the daily reporting cadence, regional demand structure, and lag-based feature design. Longer horizons are reserved for future work.

The multi-task objective combines Task 1 and Task 2 losses. Let \(\mathcal{L}_{\mathrm{demand}}\) denote a robust regression loss on the MW scale---implemented as Huber loss \cite{Huber1964RobustEstimation} averaged over the \(N\) regional demand targets---and let \(\mathcal{L}_{\mathrm{osi}} = \ell_{\mathrm{MSE}}(\hat{y}^{\mathrm{osi}}, \mathrm{OSI}(t+H))\) denote mean squared error on the bounded stress target. The joint training objective is

\[
\mathcal{L} = \mathcal{L}_{\mathrm{demand}} + \lambda_2 \, \mathcal{L}_{\mathrm{osi}},
\]

where \(\lambda_2 > 0\) weights stress learning against demand fit. Task-weight calibration and optimisation details follow in Sections~\ref{subsection:training_strategy} and~\ref{subsection:loss_function_and_optimization}.

\subsection{Dataset Description}\label{subsection:dataset_description}

We apply the multi-task formulation to historical operational records from the Bangladesh national power grid using the publicly available Bangladesh Smart Grid dataset (Mendeley Data, Version 5) \cite{Hossain2025BangladeshSmartGrid}. The daily operational timeline spans several years of system and regional reporting; key properties are summarised in Table~\ref{tab:1}.

\textbf{Data source and granularity.} The underlying archive provides one record per calendar day. Each record aggregates national generation and demand indicators together with division-level operational signals. The timeline spans 21 November 2019 to 30 December 2024, yielding 1,850 unique daily observations over a 1,866-day calendar span. Sampling is predominantly at one-day intervals; a small number of calendar gaps are present in the series and are accounted for in later modelling stages. All 45 recorded fields are complete, with no missing entries in the audited source timeline. Table~\ref{tab:1} lists the domain, regional coverage, and structural properties used in this study.

\textbf{Spatial coverage.} The spatial units in \(\mathcal{V}\) correspond to \(N = 9\) administrative divisions of Bangladesh: Barishal, Chattogram, Cumilla, Dhaka, Khulna, Mymensingh, Rajshahi, Rangpur, and Sylhet. This partitioning matches the regional reporting structure reflected in the source data and supports node-level demand forecasting across geographically distinct load centres.

\textbf{Recorded variables.} The daily records combine calendar metadata, national generation and demand scalars, exogenous operational constraints, and regional power-balance signals. Calendar fields include the date, day of week, month, year, and holiday indicators. National-level quantities document evening-peak and day-peak generation, minimum and forecast generation, and demand measured at generation and sub-station endpoints. Operational constraint fields capture gas and liquid-fuel limitations, coal supply limitations, low water levels at Kaptai Lake, and capacity under plant shutdown or maintenance. A national temperature anomaly proxy for Dhaka is also recorded. At the regional level, each division is associated with three coupled signals: evening-peak demand, supply, and load (unmet demand / shedding) in megawatts. In total, the source timeline contains 42 numeric operational variables and three non-numeric calendar descriptors alongside the date index.

\textbf{Target variables.} Two supervised targets are derived from the recorded timeline in alignment with Section~\ref{subsection:problem_definition_and_task_formulation}. For Task 1, the primary target is next-day regional evening-peak demand \(D_r(t+H)\) for each node \(r \in \mathcal{V}\), taken from the division-level demand fields. For Task 2, the Operational Stress Index is a graph-level score in \([0, 1]\) constructed from system-wide shedding intensity, generation reserve margin relative to highest available generation, and aggregate operational limitation pressure---the inputs to components \(c_1\)--\(c_3\) defined in Section~\ref{subsection:problem_definition_and_task_formulation}. Regional supply and division-level shedding are retained in the source record because they inform system context and stress composition, but only demand and OSI serve as forecasting targets under the frozen task definition.

\textbf{Dataset characteristics.} The timeline suits joint demand--stress forecasting: daily cadence matches \(H = 1\); nine divisions provide a natural node set; demand, shedding, reserve, and limitation co-exist in each record for composite stress construction; and multi-year coverage spans full seasonal cycles, holidays, and supply-constraint periods.

Prior data review documented strong demand--supply coupling, motivating demand-only Task 1 targets, and sparse regional load distributions, motivating OSI rather than per-node shedding regression. Inter-regional demand correlation supports graph-based spatio-temporal modelling in subsequent subsections.

\subsection{Data Preprocessing}\label{subsection:data_preprocessing}

The operational timeline described in Section~\ref{subsection:dataset_description} is prepared for downstream modelling through a two-stage preprocessing pipeline that precedes feature engineering. The first stage performs conservative scientific cleaning to validate and correct the source record while preserving genuine operational events. The second stage converts the cleaned timeline into a model-ready representation through chronological ordering, categorical encoding, and numeric standardisation, with all learnable transformations restricted to training data to prevent temporal leakage. No lag features, rolling statistics, graph structures, or supervised targets are created in this stage.

\textbf{Data validation and integrity assessment.} Preprocessing begins with a read-only audit of the published raw archive to establish record counts, field types, and temporal structure without altering source values. The audit confirmed 1,850 daily rows and 45 fields, with 42 numeric operational measurements and three non-numeric calendar descriptors indexed by date. Missing-value assessment across all fields returned zero absent cells. Duplicate inspection found no repeated rows and no duplicate dates. Negative measurements and implausible temperature values were absent. These checks establish that the archive is complete at the cell level and suitable for time-ordered forecasting, while deferring any modification of the raw file to the cleaning stage.

\textbf{Scientific data cleaning.} The cleaning stage ingests the published archive and produces a cleaned intermediate dataset of dimensions 1,850 $\times$ 45. Cleaning follows a conservative rule: only impossible, corrupted, duplicated, or invalid records are corrected or removed; statistically rare but operationally genuine events are retained. No rows were removed. The date field was validated and converted to a standard temporal representation without changing calendar values. Timestamp verification confirmed that all dates are parseable, unique, and strictly monotonic increasing over the range 2019-11-21 to 2024-12-30. Chronological consistency checks verified agreement between the date index and the recorded year and month fields in every row. Six inconsistent day-of-week labels were corrected deterministically from the authoritative date index; no numeric measurements were altered.

Calendar continuity was assessed against a daily index. Seventeen missing calendar days were documented in the series; these gaps were not imputed or synthetically filled during cleaning, so only observed daily records are retained. This policy preserves the authentic reporting cadence and avoids introducing artificial observations that would distort lag-based forecasting. Physical-consistency review identified regional accounting mismatches and generation-ordering discrepancies in a subset of rows; these patterns were classified as reporting or metering differences rather than impossible values and were preserved as informative operational signals. Row and column counts were unchanged between input and output, confirming that cleaning altered only invalid metadata rather than measurement content.

\textbf{Model-ready preprocessing.} The cleaned intermediate timeline serves as the sole input to the model-preparation stage. Records are sorted in ascending chronological order to enforce temporal ordering required for causal forecasting. Three categorical fields---day of week, holiday name, and holiday category---are one-hot encoded using categories observed in the training partition only, with unseen validation or test categories assigned zero contribution. This encoding supplies a fixed numeric representation of calendar and holiday context without manual category collapsing.

Forty-one numeric operational fields, comprising national generation and demand scalars, limitation indicators, temperature, calendar ordinals (year and month), and all regional demand, supply, and load measurements, are standardised by z-score scaling with parameters learned exclusively from the training partition. Validation and test partitions are transformed using the frozen training statistics only. Standardisation places heterogeneous MW-scale quantities on a common scale for subsequent learning while preventing held-out data from influencing the fitted parameters. The date index is retained as an unscaled temporal key. Only existing source fields are transformed; no new predictive attributes are introduced at this stage.

\textbf{Leakage prevention.} Temporal leakage is controlled at two levels during preprocessing. First, chronological ordering is preserved throughout cleaning and transformation so that no future observation precedes a past one in the stored timeline. Second, all learnable preprocessing parameters---the one-hot category sets and scaling means and standard deviations---are estimated on the training partition only; validation and test data receive transformation with frozen training parameters. Cleaning and model-preparation stages remain separable, so encoding and scaling cannot retroactively alter the cleaned record. Same-day stress indices and other supervision-derived quantities are not injected during this stage; the leakage policy for OSI inputs defined in Section~\ref{subsection:problem_definition_and_task_formulation} is enforced in subsequent feature construction.

\textbf{Preprocessing outputs.} The pipeline delivers a validated cleaned dataset, chronologically partitioned encoded records, and a serialised transformation pipeline freezing encoding and scaling rules for downstream feature engineering, graph construction, and multi-task training.

\subsection{Feature Engineering}\label{subsection:feature_engineering}

The leakage-aware preprocessing outputs described in Section~\ref{subsection:data_preprocessing} are extended into a structured feature set that supplies the spatio-temporal inputs defined in Section~\ref{subsection:problem_definition_and_task_formulation}. Feature engineering transforms the cleaned operational timeline and its encoded baseline representation into derived predictors that capture autoregressive load dynamics, calendar structure, grid-wide stress context, and regional heterogeneity. The design prioritises causal observability: every engineered quantity at day \(t\) is computable from information available at or before \(t\), with no reliance on future rows or held-out partition statistics. In total, 65 engineered features are constructed and merged with the preprocessed baseline by date. Raw source measurements and their preprocessed encodings are distinguished throughout from quantities constructed by deterministic transformation rules.

\textbf{Design objectives.} The feature programme serves three methodological goals aligned with the joint demand--stress formulation. First, it encodes short-horizon temporal structure appropriate to a one-day forecast horizon and a seven-day input window \(T = 7\), including autoregressive demand signals and gap-aware offsets that respect irregular calendar spacing. Second, it represents both node-local load state and national system context within each daily record, enabling the model to relate regional forecasts to grid-wide limitation and reserve conditions. Third, it enforces a strict leakage policy: same-day Operational Stress Index values are excluded from the model input specification even though an OSI-derived feature is computed for target construction and diagnostic use, consistent with Section~\ref{subsection:problem_definition_and_task_formulation}.

\textbf{Raw variables versus engineered features.} Raw variables comprise the division-level evening-peak demand, supply, and load measurements, national generation and demand scalars, operational limitation indicators, temperature, and calendar descriptors described in Section~\ref{subsection:dataset_description}. Preprocessing supplies encoded and standardised forms of these fields without creating new attributes. Engineered features are deterministic functions of raw and preprocessed quantities---lags, rolling statistics, normalised shares, grid aggregates, cyclical calendar encodings, and composite stress indicators---that are computed on the chronological timeline and merged with the baseline representation by date. Continuous engineered quantities are standardised using training-partition statistics only; a binary shedding-presence indicator is retained without scaling.

\textbf{Temporal feature engineering.} Four global temporal descriptors encode seasonality, long-run drift, and irregular sampling. Day-of-year sine and cosine features provide smooth annual periodicity without discontinuities at year boundaries. A monotonic trend index captures slow structural change across the multi-year record. An inter-observation gap feature records the elapsed days between consecutive observed records, explicitly signalling calendar discontinuities rather than treating missing days as zero-load intervals. At each node, three lag families supply autoregressive structure: one-day and seven-observation demand lags, and one-day load lags. Lags use observed-row offsets on the sorted timeline rather than fixed calendar offsets, so a seven-observation lag advances seven recorded days even when calendar gaps intervene. Seven-day demand rolling means are computed over the seven preceding observed rows with the current day excluded, yielding a past-only smoothed load level that complements point lags for short-horizon forecasting.

\textbf{Node-level engineered features.} The node input specification comprises three conceptual groups. A contemporaneous power-balance group carries processed demand, supply, and load at each division, anchoring the node to its current operational state. An autoregressive temporal group supplies demand lags at one and seven observations, a one-observation load lag, and a seven-observation demand rolling mean, encoding short-term persistence and weekly structure aligned with the input window. A regional derived group expresses each division's relative role in the national system: demand share \(D_r(t) / \sum_{j} D_j(t)\), which normalises heterogeneous scales to a common simplex representation, and load intensity \(L_r(t) / \max(D_r(t), \varepsilon)\) with \(\varepsilon = 1\) MW, which expresses unmet demand relative to local load without treating sparse shedding zeros as informative magnitude on their own. These three groups yield \(F_n = 9\) channels per region in the node tensor \(\mathbf{X}^{\mathrm{node}}\).

\textbf{Global engineered features.} The global input specification comprises six conceptual groups. A temporal and calendar group includes the four engineered descriptors above---cyclical day-of-year encodings, trend index, and inter-observation gap---together with a holiday-category summary derived by consolidating the four one-hot holiday-category channels produced in Section~\ref{subsection:data_preprocessing} into a single ordinal indicator for model input. A grid-aggregate group sums total regional demand and total regional load and computes generation reserve as the difference between highest available generation and evening-peak demand at the generation endpoint, summarising system-wide balance and headroom. An operational group aggregates gas, coal, water, and maintenance limitation indicators into total operational limitation, records whether any division reports positive shedding, and---during feature construction only---forms a composite stress score from shedding intensity, reserve margin, and limitation pressure using train-split normalisation bounds; this composite is excluded from model inputs to prevent same-day OSI leakage when forecasting \(\mathrm{OSI}(t+1)\). A weather group expresses temperature anomaly relative to train-estimated monthly means. A limitation group retains the four preprocessed operational-constraint scalars from the baseline representation. A national generation group retains two preprocessed generation-endpoint scalars. Together, these groups yield \(F_g = 17\) channels in the global tensor \(\mathbf{X}^{\mathrm{global}}\) at each timestep.

\textbf{Operational feature representation.} Operational engineering targets the constraint drivers underlying Task 2 without substituting the supervised OSI label into the input set. Total operational limitation condenses fuel, infrastructure, and maintenance constraints into a single MW-scale pressure index. Generation reserve links available capacity to contemporaneous demand. The binary any-division shedding indicator flags system-wide unmet-demand events that may precede graded stress escalation. The engineered stress composite mirrors the component structure of Section~\ref{subsection:problem_definition_and_task_formulation}---shedding ratio, reserve shortfall, and limitation pressure---but is withheld from \(\mathbf{X}^{\mathrm{global}}\) at inference time so that stress predictions at \(t+1\) cannot access a normalised stress score already computed at \(t\).

\textbf{Feature grouping.} The implemented channels are organised into eleven coalitions that separate raw regional blocks from engineered groups: regional demand, supply, and load blocks (G1--G3); engineered lags and rolling statistics (G4); regional share and intensity features (G5); calendar and trend descriptors (G6); grid aggregates (G7); the limitation stack (G8); weather anomaly (G9); national generation scalars (G10); and the shedding indicator (G11). This grouping maps each coalition to either node or global scope while preserving the distinction between contemporaneous measurements and derived predictors. The coalition definitions support grouped attribution analysis in Section~\ref{sec:results} (Table~\ref{tab:7}).

\textbf{Leakage-aware feature design.} All lag and rolling computations apply strictly backward shifts on the chronological timeline. Monthly temperature means, OSI normalisation bounds for the composite stress feature, and standardisation parameters for continuous engineered fields are estimated on the training partition only. The Operational Stress Index at day \(t\) is never included among the seventeen global model inputs. Demand and load histories through day \(t\) remain admissible for forecasting at \(t+1\). Calendar gaps are propagated through the gap-day feature and gap-aware lags rather than imputed.

\textbf{Final feature representation.} At each forecast origin, the model receives a lookback window of \(T = 7\) days. Node inputs form \(\mathbf{X}^{\mathrm{node}} \in \mathbb{R}^{T \times N \times F_n}\) with \(N = 9\) and \(F_n = 9\); global inputs form \(\mathbf{X}^{\mathrm{global}} \in \mathbb{R}^{T \times F_g}\) with \(F_g = 17\), instantiating the input tensors defined in Section~\ref{subsection:problem_definition_and_task_formulation}.

\subsection{Graph Construction}\label{subsection:graph_construction}

Feature tensors encode temporal signals but not spatial coupling among regional loads. Bangladesh evening-peak demand exhibits strong inter-division correlation from shared national growth, seasonality, and calendar shocks. We adopt a graph with each division as a vertex and weighted edges encoding historical demand co-movement, yielding a fixed spatial prior at every forecast origin.

\textbf{Node definition.} The vertex set \(\mathcal{V}\) contains \(N = 9\) nodes, one per administrative division---Barishal, Chattogram, Cumilla, Dhaka, Khulna, Mymensingh, Rajshahi, Rangpur, and Sylhet---in fixed alphabetical order for reproducible indexing. Each node is aligned with the corresponding row of the node feature tensor \(\mathbf{X}^{\mathrm{node}}\) described in Section~\ref{subsection:feature_engineering}: contemporaneous demand, supply, and load; autoregressive lags and rolling means; and regional share and intensity descriptors. National and system-level quantities remain global context in \(\mathbf{X}^{\mathrm{global}}\) rather than separate graph vertices, preserving a clear separation between node-local load state and grid-wide conditioning.

\textbf{Spatial dependency modelling.} Spatial dependence is modelled through undirected edges whose weights reflect pairwise statistical association in regional demand. During graph design, three construction strategies were specified---geographical adjacency, correlation-thresholded adjacency, and a hybrid of both---so that domain priors and data-driven coupling could be compared under a common node set. For the proposed framework, the correlation-only strategy is adopted as the definitive edge rule; alternative formulations are described in a later subsection, and the correlation-graph construction steps are detailed below.

\textbf{Correlation computation.} Pairwise edge weights are derived from Pearson correlation coefficients \(\rho_{ij}\) computed on regional evening-peak demand series using training-partition observations only (\(n = 1{,}295\) daily records). Restricting correlation estimation to the training partition prevents held-out periods from influencing the spatial prior. Demand was selected as the correlation basis because it is the primary Task 1 target and exhibits the strongest and most stable inter-regional association in the exploratory record. Train-period demand correlations are uniformly positive across division pairs, with the weakest associations confined to a small subset of pairs; most divisions co-vary strongly over shared seasonal and growth cycles rather than evolving independently.

\textbf{Edge definition and sparsification.} An undirected edge \((i, j)\) is retained when \(\rho_{ij} \geq \tau\), with correlation threshold \(\tau = 0.65\). Edge weight equals \(\rho_{ij}\) for retained pairs and zero otherwise. The threshold follows the correlation cut-off adopted during exploratory analysis and graph design: it retains pairs with substantively positive demand co-movement, excludes the weakest associations, and yields a connected graph suitable for spatial message passing. Under this rule, 33 of 36 possible undirected pairs are retained (91.7\% pair density). Self-loops are excluded by setting diagonal entries to zero, so each node does not receive a trivial self-message through the spatial prior. The resulting topology is symmetric, non-negative, and connected.

\textbf{Graph normalisation.} Raw correlation weights are row-normalised to produce the adjacency matrix \(\mathbf{A}\) used in the proposed framework. Each row is divided by its sum so that outgoing edge weights from a node form a convex combination summing to unity. From a methodological standpoint, row normalisation expresses each division's spatial neighbourhood as a weighted mixture of its neighbours rather than an unscaled sum of correlation magnitudes, which stabilises cross-division aggregation when node degree and edge strength vary. It also prevents high-degree nodes from dominating neighbour contributions solely because they retain more edges. The normalised matrix is fixed after training-partition estimation and applied unchanged across validation and test periods.

\textbf{Static versus dynamic graph.} The adjacency matrix is static: edge structure and weights are estimated once from the full training timeline and held fixed thereafter. Time-varying graph refresh---for example, through rolling correlation windows---was considered during graph design but deferred to keep the spatial prior reproducible and leakage-safe. Dynamic national context instead enters through the global feature channels and the temporal dimension of \(\mathbf{X}^{\mathrm{node}}\), while \(\mathbf{A}\) supplies a stable map of historical inter-division demand coupling.

\textbf{Alternative graph formulations.} The geographical graph connects border-adjacent divisions with uniform binary edges (21 undirected pairs; 58.3\% density), row-normalised for aggregation. This variant encodes an administrative-border prior on plausible power-exchange neighbourhoods but assigns equal weight to all geographic links regardless of observed coupling strength. The hybrid graph retains an edge when either a geographical link exists or \(\rho_{ij} \geq 0.85\), with weight \(\rho_{ij}\) (24 undirected pairs; 66.7\% density). This variant augments the geographic skeleton with the strongest non-border demand associations. Neither alternative enters the proposed framework, which employs the correlation-only rule with \(\tau = 0.65\) described above.

\textbf{Final graph representation.} Graph construction yields row-normalised adjacency \(\mathbf{A} \in \mathbb{R}^{N \times N}\) and a derived attention-bias matrix (log-scaled positive entries, masked absent edges) for additive spatial masking. Both matrices are fixed, train-estimated artefacts aligned to alphabetical node order and accompany each forecast window.

\subsection{Proposed STGT Architecture}\label{subsection:proposed_stgt_architecture}

The Correlation-Aware Multi-Task Forecasting Framework---implemented as Parallel-Fusion Spatio-Temporal Graph Transformer (PF-STGT)---integrates graph structure and feature tensors in one model that jointly forecasts next-day regional demand and OSI (Fig.~\ref{fig:1}). A parallel dual-branch design combines graph-aware spatial attention for inter-division coupling with a temporal transformer for within-division dynamics; learnable gating fuses the branches before task-specific heads.

\textbf{Model inputs and representational layers.} Three input layers feed the architecture, and each serves a distinct role. The data representation comprises the node tensor \(\mathbf{X}^{\mathrm{node}} \in \mathbb{R}^{T \times N \times F_n}\) and the global tensor \(\mathbf{X}^{\mathrm{global}} \in \mathbb{R}^{T \times F_g}\), constructed as described in Section~\ref{subsection:feature_engineering} for a lookback window of \(T = 7\) days ending at forecast origin \(t\). The graph representation comprises the row-normalised adjacency matrix \(\mathbf{A}\) and the derived attention-bias matrix from Section~\ref{subsection:graph_construction}; these artefacts are fixed across forward passes and do not vary with the input window. The neural architecture maps embedded inputs through spatial and temporal encoders, fuses the resulting hidden states, and decodes task outputs. Supervision targets---regional demand at \(t+1\) and \(\mathrm{OSI}(t+1)\)---are not model inputs; the leakage policy in Section~\ref{subsection:problem_definition_and_task_formulation} excludes same-day OSI from the feature set when forecasting next-day stress.

\textbf{Input embedding.} Node and global tensors are projected into a shared latent space \(d_{\mathrm{model}}\). Node channels receive a learnable regional identity embedding; global channels broadcast national calendar, limitation, and generation context at each timestep; positional encoding marks lookback position. The resulting \(\mathbf{H}_0 \in \mathbb{R}^{T \times N \times d_{\mathrm{model}}}\) feeds both encoder branches.

\textbf{Graph integration.} The correlation graph enters the spatial encoder through an additive attention bias derived from \(\mathbf{A}\). For each ordered node pair \((i,j)\), positive adjacency weights contribute a log-scaled bias term to the attention score, while absent edges receive a large negative mask so that message passing is restricted to the correlation-supported neighbourhood. This mechanism couples learned attention patterns with the train-estimated spatial prior from Section~\ref{subsection:graph_construction} rather than relying on free-form pairwise weights alone. Graph integration is applied independently at each of the \(T\) timesteps in the spatial branch, allowing inter-division information exchange to vary day by day within the window while the underlying adjacency structure remains static.

\textbf{Spatial encoder.} The spatial branch is a graph transformer \cite{Vaswani2017Attention,Kipf2017GCN} that applies multi-head self-attention over the \(N\) nodes at each timestep. Given hidden states at timestep \(t\), attention scores combine scaled query--key inner products with the graph bias matrix, and the softmax-normalised weights aggregate value vectors across nodes:

\[
\mathrm{Attn}_{\mathrm{spatial}}(\mathbf{Q}, \mathbf{K}, \mathbf{V}) = \mathrm{softmax}\!\left(\frac{\mathbf{Q}\mathbf{K}^\top}{\sqrt{d}} + \mathbf{B}_{\mathrm{adj}}\right)\mathbf{V},
\]

where \(\mathbf{B}_{\mathrm{adj}}\) denotes the attention-bias matrix derived from \(\mathbf{A}\). Stacked encoder layers with residual connections and position-wise feed-forward transformations refine \(\mathbf{H}_0\) into a spatially encoded sequence \(\mathbf{H}_{\mathrm{spatial}} \in \mathbb{R}^{T \times N \times d_{\mathrm{model}}}\). A graph transformer was selected over fixed-weight graph convolutions because division coupling strength varies across pairs and because attention weights provide an interpretable record of spatial information flow. The spatial encoder addresses the limitation of treating regions as independent: it propagates demand-relevant signals along correlation-weighted edges at each day in the lookback window.

\textbf{Temporal encoder.} The temporal branch is a transformer encoder \cite{Vaswani2017Attention} applied independently to each node's \(T\)-step sequence, with parameters shared across nodes for efficiency. At each node, multi-head self-attention operates over timesteps, capturing short-range autoregressive structure and weekly periodicity aligned with the seven-day input window and lag-based features in Section~\ref{subsection:feature_engineering}. The encoder outputs \(\mathbf{H}_{\mathrm{temporal}} \in \mathbb{R}^{T \times N \times d_{\mathrm{model}}}\) with the same shape as the spatial branch. A full transformer encoder was preferred over a single shallow attention layer because the operational record exhibits multi-day seasonal structure within the window. Causal masking is not applied: the window contains only observations at or before the forecast origin, so ex-post encoding of the full lookback is consistent with the day-ahead forecasting setting.

\textbf{Feature fusion.} Spatial and temporal branches encode complementary aspects of the input---cross-division coupling and within-division dynamics---and are integrated through parallel fusion rather than a strict sequential pipeline. A learnable gate computed from the concatenated branch outputs forms a convex combination at each timestep and node:

\[
\mathbf{H}_{\mathrm{fused}} = \mathbf{G} \odot \mathbf{H}_{\mathrm{spatial}} + (\mathbf{1} - \mathbf{G}) \odot \mathbf{H}_{\mathrm{temporal}},
\]

where \(\mathbf{G} = \sigma(\mathrm{Linear}([\mathbf{H}_{\mathrm{spatial}}; \mathbf{H}_{\mathrm{temporal}}]))\) and \(\odot\) denotes element-wise multiplication. Parallel fusion allows the model to emphasise spatial or temporal evidence depending on context---for example, synchronised national load movement versus division-specific short-term persistence---without forcing one inductive bias to precede the other. Sequential spatial-then-temporal or temporal-then-spatial designs were rejected during architecture design because they delay one form of coupling and can blur the temporal structure before cross-division mixing, or defer spatial propagation of same-day regional shocks.

\textbf{Shared latent representation.} The fused sequence is reduced to a per-node shared embedding by selecting the last observed timestep in the window:

\[
\mathbf{H}_{\mathrm{shared}} = \mathbf{H}_{\mathrm{fused}}[:, T, :, :] \in \mathbb{R}^{N \times d_{\mathrm{model}}}.
\]

This tensor constitutes the common latent state from which both tasks are decoded. A graph-level readout vector is additionally formed by mean-pooling \(\mathbf{H}_{\mathrm{shared}}\) across nodes and concatenating the result with the embedded global context at the final timestep, yielding a compact system-wide summary for stress prediction. The shared representation enforces multi-task information sharing: demand and OSI heads draw on the same spatio-temporal encoding of the input window rather than independent models that would duplicate feature extraction.

\textbf{Demand prediction head.} Task 1 is decoded through a dedicated per-node linear head that maps each row of \(\mathbf{H}_{\mathrm{shared}}\) to a scalar next-day demand forecast in megawatts, producing the regional output vector \(\hat{\mathbf{y}}^{\mathrm{demand}} \in \mathbb{R}^N\) defined in Section~\ref{subsection:problem_definition_and_task_formulation}. A node-specific linear mapping is sufficient because \(\mathbf{H}_{\mathrm{shared}}\) already encodes division identity through the regional embedding and spatial attention; the head's role is to project the shared latent state onto the MW scale for each region. This design yields nine heterogeneous forecasts from one trunk while preserving parameter efficiency relative to nine independent sequence models.

\textbf{OSI prediction head.} Task 2 is decoded through a graph-level head that consumes three complementary views of the shared state: the mean-pooled graph summary concatenated with the final-timestep global embedding, and the flattened node-level shared representation across all divisions. A feed-forward network with a bounded output activation maps this combined vector to a single scalar \(\hat{y}^{\mathrm{osi}} \in [0, 1]\). The OSI head requires both aggregated system context and division-resolved latent states because operational stress in this formulation is a national score influenced by spatially distributed load and constraint patterns. Restricting the stress decoder to graph-pooled information alone would discard node-level variation retained in \(\mathbf{H}_{\mathrm{shared}}\); restricting it to a single node would fail to represent system-wide stress.

\textbf{Multi-task learning mechanism.} Demand and OSI share embedding, both encoders, fusion, and \(\mathbf{H}_{\mathrm{shared}}\); only decoding layers differ. Shared encoding exploits common calendar, limitation, and load history while preserving distinct output semantics (node-resolved MW demand versus graph-resolved bounded OSI). Loss balancing is specified in Section~\ref{subsection:loss_function_and_optimization}.

\textbf{Architectural rationale.} PF-STGT addresses three requirements: correlation-only graph conditioning matched to strong inter-regional demand coupling; parallel spatial and temporal transformer branches for cross-division and within-division dynamics; and a shared latent core with specialised demand and stress heads for joint day-ahead forecasting from one spatio-temporal pass.

\subsection{Training Strategy}\label{subsection:training_strategy}

PF-STGT is trained on chronologically partitioned samples under the leakage policies of Sections~\ref{subsection:problem_definition_and_task_formulation} and~\ref{subsection:data_preprocessing}. Future observations never influence past inputs; preprocessing and graph parameters are frozen from training data; model selection uses validation only. Sliding windows with aligned multi-task targets convert the feature timeline and static graph into supervised examples without crossing partition boundaries.

\textbf{Chronological partition.} The operational timeline is divided into training, validation, and test subsets by ascending date, following the chronological split applied during preprocessing. The training partition spans 21 November 2019 to 15 June 2023 (1,295 daily records); validation spans 16 June 2023 to 19 March 2024 (277 records); and test spans 20 March 2024 to 30 December 2024 (278 records). Partition boundaries are strictly non-overlapping: the latest training date precedes the earliest validation date, and the latest validation date precedes the earliest test date. Chronological partitioning is required because random shuffling of days would place future grid states in the training set and inflate performance estimates. The split ratios and date ranges are summarised in Table~\ref{tab:1}.

\textbf{Sequence generation and warm-up exclusion.} Training examples are generated independently within each partition. For a partition containing \(R\) chronologically ordered daily records, a valid forecast origin must satisfy two constraints. First, the lookback window of length \(T = 7\) must lie entirely within the partition and contain only rows with complete engineered features. Second, the supervision index must lie one horizon step (\(H = 1\)) ahead of the window end and remain inside the same partition. The first seven rows of each partition are excluded as warm-up because seven-observation demand lags and seven-day rolling means are undefined at the series start; this policy aligns with the lag-based feature design in Section~\ref{subsection:feature_engineering}. The resulting numbers of valid sliding-window samples per partition are reported in Table~\ref{tab:1}. Calendar gaps in the source timeline are preserved: windows span consecutive observed rows within a partition rather than imputed calendar days.

\textbf{Sliding-window construction.} Each training example is built by sliding a fixed-length window one observed row at a time. For forecast origin index \(t_{\mathrm{end}}\), the input window covers days \([t_{\mathrm{end}} - T + 1, \, t_{\mathrm{end}}]\) and supplies the node tensor \(\mathbf{X}^{\mathrm{node}}\) and global tensor \(\mathbf{X}^{\mathrm{global}}\) described in Section~\ref{subsection:feature_engineering}. The static adjacency matrix \(\mathbf{A}\) and its derived attention-bias matrix from Section~\ref{subsection:graph_construction} are attached to every sample because the graph prior does not vary across origins. Window construction is identical across partitions; only the allowable range of \(t_{\mathrm{end}}\) changes with partition size and boundary position. Sliding-window enumeration maximises the number of supervised examples available within each period without altering the one-day forecast horizon.

\textbf{Multi-task target preparation.} Supervision is prepared separately from model inputs. For Task 1, regional evening-peak demand at day \(t_{\mathrm{end}} + H\) is extracted from the cleaned interim timeline in raw megawatts for each of the nine divisions, producing the target vector \(\mathbf{y}^{\mathrm{demand}} \in \mathbb{R}^N\). For Task 2, the Operational Stress Index at the same future day is computed from the component formulae in Section~\ref{subsection:problem_definition_and_task_formulation} using train-partition min--max bounds for normalisation; the resulting scalar \(\mathrm{OSI}(t_{\mathrm{end}} + H)\) lies in \([0, 1]\). Demand targets use unscaled MW values from the clean operational record so that the demand head learns on a physically interpretable scale. OSI component bounds are fit once on the training partition and frozen, consistent with the leakage policy that forbids using validation or test statistics in target construction. Same-day OSI at \(t_{\mathrm{end}}\) is never included among input features when predicting \(\mathrm{OSI}(t_{\mathrm{end}} + H)\).

\textbf{Batch formation.} Training iterates over mini-batches of windowed samples drawn from the training partition. Within a batch, node and global feature tensors are stacked along the batch dimension; demand and OSI targets are stacked correspondingly. The adjacency and attention-bias matrices are shared across all samples in a batch because the graph is static. Training batches are shuffled each epoch to vary the order of independent window samples; this shuffling does not alter the chronological ordering of observations within any individual window. Validation and test loaders preserve chronological sample order so that temporal structure remains inspectable. Batch construction therefore separates variable per-origin features from fixed graph artefacts shared across the national grid.

\textbf{Validation protocol.} During training, model parameters are updated using only the training partition. After each epoch, performance is assessed on the validation partition using the same windowing and target rules, without parameter updates. Validation metrics include macro demand mean absolute error over the nine regions and stress mean absolute error for the graph-level OSI target. For the multi-task PF-STGT, checkpoint selection is guided by a balanced composite validation score that reflects both demand and stress prediction error, so that model selection does not optimise a single task at the expense of the other. The test partition is not used during training, hyperparameter selection, or early stopping; it is reserved for the held-out evaluation reported in Section~\ref{sec:results}.

\textbf{Early stopping and checkpoint selection.} Training stops when the composite validation score fails to improve for a fixed patience interval. The checkpoint with the lowest composite validation score is retained for test evaluation.

\textbf{Reproducibility considerations.} A fixed random seed governs initialisation and shuffling; frozen datasets, features, and graph specifications are verified before training; and partition boundaries, window length, horizon, and warm-up rules are applied identically across loaders. Table~\ref{tab:2} summarises frozen hyperparameters and the multi-task loss specification.

\subsection{Loss Function and Optimization}\label{subsection:loss_function_and_optimization}

This subsection specifies the objective function and optimisation procedure. Minimisation uses a weighted multi-task loss over training mini-batches. Demand targets remain in megawatts at evaluation; the training objective applies scale-aware transformation so demand and stress gradients remain commensurate. Updates use Adam with weight decay, ReduceLROnPlateau scheduling, and gradient-norm clipping.

\textbf{Multi-task objective.} Let \(\hat{\mathbf{y}}^{\mathrm{demand}} \in \mathbb{R}^N\) denote the vector of predicted regional evening-peak demands and \(\hat{y}^{\mathrm{osi}} \in [0,1]\) the predicted Operational Stress Index for a single forecast origin, with corresponding targets \(\mathbf{y}^{\mathrm{demand}}\) (MW) and \(y^{\mathrm{osi}}\) prepared as in Section~\ref{subsection:training_strategy}. The joint training objective extends the formulation introduced in Section~\ref{subsection:problem_definition_and_task_formulation} by separating raw demand fit from stress fit and introducing an explicit demand-scale normalisation before task weighting:

\[
\mathcal{L} = \frac{\mathcal{L}_{\mathrm{demand}}^{\mathrm{raw}}}{100} + \lambda_2 \, \mathcal{L}_{\mathrm{osi}},
\]

where \(\lambda_2 = 20\) is the frozen stress-task weight and the divisor \(100\) (MW) maps the demand loss term to a magnitude comparable with the bounded OSI loss. Minimisation is performed over all trainable parameters \(\boldsymbol{\theta}\) of the PF-STGT model by mini-batch optimisation using the Adam optimiser \cite{Kingma2015Adam} with batch size \(B = 32\).

\textbf{Demand loss.} Regional demand errors are aggregated with a macro-averaged Huber loss over the \(N = 9\) divisions:

\[
\mathcal{L}_{\mathrm{demand}}^{\mathrm{raw}} = \frac{1}{N} \sum_{r=1}^{N} \mathrm{Huber}_{\delta}\!\left(\hat{y}^{\mathrm{demand}}_r,\, y^{\mathrm{demand}}_r\right),
\]

where \(\mathrm{Huber}_{\delta}(a, b)\) is the element-wise Huber function with threshold \(\delta = 1.0\) MW:

\[
\mathrm{Huber}_{\delta}(a, b) =
\begin{cases}
\frac{1}{2}(a - b)^2, & |a - b| \le \delta, \\
\delta\left(|a - b| - \frac{\delta}{2}\right), & \text{otherwise}.
\end{cases}
\]

Huber regression was selected in preference to pure squared error because evening-peak demand exhibits occasional upper-tail excursions; the quadratic region penalises moderate errors smoothly, while the linear region limits the influence of large outliers that would otherwise dominate batch gradients. Macro-averaging across regions assigns equal weight to each division in the loss, consistent with the node-level forecasting objective in Section~\ref{subsection:problem_definition_and_task_formulation}. Targets remain in raw megawatts during training so that the demand head learns on the operational scale used for evaluation; only the aggregated loss contribution entering \(\mathcal{L}\) is divided by \(100\) MW.

\textbf{OSI loss.} The stress head output passes through a sigmoid activation so that \(\hat{y}^{\mathrm{osi}}\) lies in \([0, 1]\), matching the bounded OSI target constructed in Section~\ref{subsection:problem_definition_and_task_formulation}. Stress fit is measured with mean squared error:

\[
\mathcal{L}_{\mathrm{osi}} = \frac{1}{B}\sum_{i=1}^{B}\left(\hat{y}^{\mathrm{osi}}_i - y^{\mathrm{osi}}_i\right)^2
\]

at batch level (equivalently, MSE over scalar OSI predictions in each mini-batch). Squared error on a unit-interval target provides a smooth, well-conditioned regression signal for the graph-level stress scalar and aligns with the continuous OSI formulation adopted over binary or quantised stress labels.

\textbf{Task weighting and scale balancing.} Without rescaling, \(\mathcal{L}_{\mathrm{demand}}^{\mathrm{raw}}\) typically attains values on the order of tens of megawatts, whereas \(\mathcal{L}_{\mathrm{osi}}\) remains close to zero because OSI targets are normalised to \([0, 1]\). A direct sum of the two terms would therefore allocate almost all gradient magnitude to the demand head, leaving the stress head under-trained despite joint architectural sharing. Dividing \(\mathcal{L}_{\mathrm{demand}}^{\mathrm{raw}}\) by \(100\) MW compresses the demand term to a \(\mathcal{O}(1)\) scale appropriate for combination with MSE on OSI. The weight \(\lambda_2 = 20\) further elevates the stress contribution so that both task heads receive comparable optimisation pressure while preserving megawatt-scale demand targets. The composite validation score used for early stopping in Section~\ref{subsection:training_strategy} applies the same \(100\) MW scaling to demand error, keeping model selection consistent with the training objective. The chosen \(\lambda_2\) and normalisation divisor are fixed prior to training and are not tuned on the test partition.

\textbf{Optimizer and weight decay.} Parameters are updated with the Adam optimiser \cite{Kingma2015Adam} at learning rate \(\eta = 5 \times 10^{-4}\) and weight decay \(\lambda_{\mathrm{wd}} = 10^{-4}\). Adam provides per-parameter adaptive step sizes that accommodate the heterogeneous gradient magnitudes arising from the parallel spatial and temporal transformer branches and the two task-specific heads. Weight decay supplies mild \(\ell_2\) regularisation on trainable weights, discouraging overfitting on the finite daily training window without altering the loss formulation explicitly. Optimisation runs for up to \(200\) epochs, subject to the early-stopping rule in Section~\ref{subsection:training_strategy}.

\textbf{Learning-rate schedule.} After each training epoch, the learning rate is adjusted by a ReduceLROnPlateau scheduler that monitors validation macro demand MAE. When this metric fails to improve for five consecutive epochs, the learning rate is multiplied by a factor of \(0.5\). Validation demand MAE is preferred for learning-rate adaptation because it varies smoothly across epochs and supplies a stable optimisation signal on the chronologically ordered validation set; model selection and early stopping nonetheless rely on the composite validation score defined in Section~\ref{subsection:training_strategy}, so checkpoint retention reflects balanced multi-task performance rather than demand fit alone. The scheduler therefore operates independently of the early-stopping criterion, allowing learning-rate decay to proceed even when the composite score has not yet triggered termination.

\textbf{Gradient clipping.} Global gradient-norm clipping with maximum norm \(1.0\) is applied immediately before each optimiser step. Clipping bounds the \(\ell_2\) norm of the concatenated parameter gradient vector, mitigating occasional large backpropagation steps that can arise from outlier demand residuals or from attention layers in the graph transformer branch. This stabilises training across epochs without changing the defined loss function.

\textbf{Numerical precision.} Full-precision arithmetic is used throughout training. This methodological choice prioritises numerical stability on the modest daily sample size over the throughput gains of reduced-precision optimisation, which was not required for reliable convergence of the PF-STGT model.

\textbf{Optimisation rationale.} Huber macro-averaged demand loss limits outlier sensitivity; demand-loss normalisation and \(\lambda_2\) prevent collapse to demand-only learning; and Adam, weight decay, plateau scheduling, gradient clipping, and full-precision arithmetic support stable convergence on fewer than \(1{,}300\) training windows. Frozen hyperparameters appear in Table~\ref{tab:2}.


% --- Section floats (publication package) ---
\input{tables/table_01_dataset.tex}
\input{tables/table_02_training.tex}
\input{figures/figure_01.tex}
\input{figures/figure_02.tex}
