\section{Introduction}\label{sec:introduction}

National grid operators must anticipate regional load and assess system pressure on a day-ahead horizon to coordinate generation dispatch, reserve planning, and constraint management. In developing economies, supply limitations and load shedding often coincide with rapid demand growth, so planners need forecasts that show both where load will concentrate and how severely operational constraints may bind system-wide. On the Bangladesh national power system, operational data are reported across nine administrative divisions with regional evening-peak demand and limitation dynamics co-visible in each daily record \cite{Hossain2025BangladeshSmartGrid}. Effective planning therefore depends on spatially disaggregated forecasts that remain temporally aligned with system-wide stress indicators.

Despite the availability of such data, forecasting practice often treats load prediction and stress monitoring as separate pipelines. Classical and ensemble models \cite{Breiman2001RandomForest,Chen2016XGBoost,Musbah2025LoadForecasting} forecast demand without explicit inter-division graph structure; shallow recurrent or graph-convolution architectures \cite{Hochreiter1997LSTM,Zhao2020TGCN} may fail to combine long-range temporal context with informative multi-node representations. Graph-based methods from residential or microgrid settings often rely on geographical adjacency that may not capture data-driven co-movement among regional loads \cite{Kipf2017GCN}. Operational stress---from reserve shortfall, limitation pressure, and shedding intensity---is seldom forecast jointly with regional demand; shedding studies \cite{Paphitis2025SpatialLoadShedding} typically address control rather than day-ahead graded stress indices. Explainability is also rarely integrated with multi-node graph-temporal models that emit coupled load and stress outputs \cite{Baur2024XAIReview,Lundberg2017SHAP}.

We address this gap by formulating day-ahead forecasting as a supervised multi-task problem \cite{Caruana1997Multitask} over nine spatial units. Task~1 predicts next-day regional evening-peak demand (MW); Task~2 predicts a graph-level Operational Stress Index (OSI) in $[0,1]$ summarising system-wide pressure from shedding intensity, generation reserve margin, and operational limitation components. Demand and OSI are operationally related yet statistically non-redundant: demand describes expected load by region, whereas OSI encodes constraint-related pressure not fully determined by demand alone. The central research question is whether a single spatio-temporal framework can learn coupled regional demand and operational stress forecasts from historical Bangladesh grid records under chronological evaluation.

This paper proposes the Correlation-Aware Multi-Task Forecasting Framework, implemented as a Parallel-Fusion Spatio-Temporal Graph Transformer (PF-STGT). The framework ingests a seven-day lookback of node-level and national context features with a train-estimated correlation graph, combines graph-aware spatial attention with temporal transformer capacity \cite{Vaswani2017Attention}, and decodes regional demand and bounded OSI through task-specific heads with a repaired multi-task loss. Evaluation on chronologically partitioned Bangladesh grid records includes classical, recurrent, and graph-convolution benchmarks, configuration ablations, inferential testing, and post-hoc attribution.

The principal contributions are as follows. First, a correlation-graph PF-STGT for joint day-ahead regional demand and OSI prediction on Bangladesh national grid data. Second, evidence-driven architecture selection favouring correlation-only adjacency over geographical or hybrid priors. Third, a repaired multi-task training protocol with explicit demand--stress loss balancing. Fourth, an empirical evaluation programme with benchmarks, ablations, Wilcoxon testing with Bonferroni correction, and bootstrap confidence intervals. Fifth, integrated explainability with multi-method attribution for both tasks.

Section~\ref{sec:methodology} presents the methodology; Sections~\ref{sec:experimental_setup}--\ref{sec:results} report experiments and results; Sections~\ref{sec:discussion}--\ref{sec:conclusion} discuss findings and conclude.
