\section{Methodology}\label{sec:methodology}

\subsection{Problem Formulation}

Let $\mathcal{V} = \{1, \ldots, N\}$ denote $N{=}9$ administrative divisions. At forecast origin $t$, the model receives a lookback window of $T{=}7$ days and predicts targets at horizon $H{=}1$. Node features $\mathbf{X}^{\mathrm{node}} \in \mathbb{R}^{T \times N \times F_n}$ ($F_n{=}9$) encode division-level demand, supply, load, lags, rolling means, and regional shares. Global features $\mathbf{X}^{\mathrm{global}} \in \mathbb{R}^{T \times F_g}$ ($F_g{=}17$) encode calendar indicators, grid aggregates, limitation variables, and national generation scalars. Task~1 outputs regional demand $\hat{\mathbf{y}}^{\mathrm{demand}} \in \mathbb{R}^N$ (MW); Task~2 outputs graph-level OSI $\hat{y}^{\mathrm{osi}} \in [0,1]$.

OSI integrates three train-normalised components:
\begin{align}
c_1 &= L_{\mathrm{total}}/D_{\mathrm{total}}, \quad
c_2 = 1 - GR/\mathrm{Highest\_Gen}, \quad
c_3 = \mathrm{TOL}/\mathrm{Highest\_Gen}, \nonumber\\
\mathrm{OSI} &= \mathrm{mean}\bigl(\mathrm{minmax}_{\mathrm{train}}(c_1),\,
\mathrm{minmax}_{\mathrm{train}}(c_2),\,
\mathrm{minmax}_{\mathrm{train}}(c_3)\bigr),
\end{align}
where $L_{\mathrm{total}}$, $D_{\mathrm{total}}$, $GR$, and $\mathrm{TOL}$ denote total shedding, demand, generation reserve, and operational limitation. Same-day OSI is excluded from inputs when forecasting $\mathrm{OSI}(t{+}1)$ to prevent leakage.

The joint objective combines macro-averaged Huber loss on demand \cite{Huber1964RobustEstimation} with MSE on OSI:
\begin{equation}
\mathcal{L} = \mathcal{L}_{\mathrm{demand}}^{\mathrm{raw}}/100 + \lambda_2 \, \mathcal{L}_{\mathrm{osi}}, \quad \lambda_2 = 20.
\label{eq:loss}
\end{equation}
Demand-loss scaling prevents collapse to demand-only learning when OSI targets lie on $[0,1]$.

\subsection{Dataset and Graph Construction}

We use the Bangladesh Smart Grid dataset (Mendeley Data, Version~5) \cite{Hossain2025BangladeshSmartGrid}: 1,850 daily records (Nov.\ 2019--Dec.\ 2024) across nine divisions. Chronological partitions yield 1,281 training, 263 validation, and 264 test sliding windows (Table~\ref{tab:setup}). All preprocessing parameters---scaling, encoding, graph weights, OSI bounds---are estimated on training data only.

Pairwise Pearson correlations $\rho_{ij}$ on training-partition evening-peak demand define undirected edges with $\rho_{ij} \geq 0.65$; weights equal $\rho_{ij}$ and are row-normalised to form adjacency $\mathbf{A}$. This correlation-only graph (33 of 36 pairs retained) replaces geographical and hybrid priors following ablation evidence. Geographical adjacency (21 pairs) and hybrid adjacency (24 pairs) serve as ablation comparators only.

\subsection{PF-STGT Architecture}

Fig.~\ref{fig:1} shows the end-to-end pipeline; Fig.~\ref{fig:2} details configuration S2. Node and global tensors are embedded with regional identity and positional encoding. A \emph{spatial branch} applies graph-transformer self-attention over nodes at each timestep with additive bias from $\mathbf{A}$:
\begin{equation}
\mathrm{Attn}_{\mathrm{spatial}}(\mathbf{Q}, \mathbf{K}, \mathbf{V}) =
\mathrm{softmax}\!\left(\frac{\mathbf{Q}\mathbf{K}^\top}{\sqrt{d}} + \mathbf{B}_{\mathrm{adj}}\right)\mathbf{V}.
\label{eq:spatial}
\end{equation}
A \emph{temporal branch} applies a shared transformer encoder over each node's seven-day sequence. Learnable gating fuses the branches; the final timestep yields shared node embeddings $\mathbf{H}_{\mathrm{shared}}$. A per-node linear head decodes regional demand; a graph-level head combining pooled node states and global context decodes OSI through a bounded activation. Both tasks share encoders and differ only in decoding layers.

\input{tables/table_01_setup}
\input{figures/figure_01.tex}
\input{figures/figure_02.tex}
