\section{Discussion}\label{sec:discussion}

These results demonstrate that one correlation-graph PF-STGT can supply operationally coupled day-ahead load and stress visibility from shared representations on Bangladesh national grid data. The discussion below interprets \emph{why} the observed patterns arise, what they imply for the Bangladesh setting, and how they should inform deployment decisions.

\subsection{Why Correlation Graphs Outperform Geography}

Geographical adjacency encodes plausible power-exchange neighbourhoods but assigns equal weight to border links regardless of observed demand coupling. On the Bangladesh national footprint, train-period evening-peak demand correlations are uniformly positive across divisions, reflecting shared national growth trajectories, holiday calendars, and temperature-driven seasonality that propagate system-wide rather than along administrative borders. The correlation graph at $\tau{=}0.65$ retains 33 of 36 pairs because most divisions co-vary strongly; geographical adjacency retains only 21 border-adjacent pairs and therefore under-connects divisions that move together without sharing a border.

A5 (geographical graph) attains marginally lower OSI MAE (0.0340) yet 9.33~MW higher demand error than S2, confirming that stress-optimal and demand-optimal topologies diverge. Geographical conditioning significantly degrades demand ($p{=}1.48{\times}10^{-4}$) because it restricts spatial message passing to a sparse, equal-weight skeleton that does not reflect the dense co-movement structure the transformer encoder exploits. Correlation weighting, by contrast, supplies graded edge strengths aligned with predictive coupling, which the graph-attention bias in Eq.~\eqref{eq:spatial} can refine further at each timestep. This pattern is consistent with developing-economy national grids where inter-regional synchronisation dominates border-local exchange \cite{Zhao2024GraphAttentionTransformer,Bentsen2023GraphTransformer}.

\subsection{Why Multi-Task Learning Behaves as Observed}

Demand and OSI share calendar structure, limitation dynamics, and national generation context but differ in output semantics: regional megawatt levels versus a bounded system-wide stress scalar. Shared spatio-temporal encoding allows the model to learn representations of limitation-relevant structure---visible in attribution rankings for coalition G8 (limitation stack) and G6 (calendar/trend)---that single-task demand models must infer indirectly through global covariates alone. Without repaired loss balancing, megawatt-scale Huber gradients suppress OSI head training; dividing demand loss by 100~MW and setting $\lambda_2{=}20$ restores gradient parity so both heads receive comparable optimisation pressure.

The demand-only ceiling (A4, 86.89~MW) is 1.76~MW below S2 because the stress head introduces a regularising signal that slightly offsets pure demand fitting, while joint training enables OSI outputs with $R^2{=}0.745$. This is not task-interference in the destructive sense but a documented Pareto trade-off: organisations requiring load forecasts alone may prefer A4; centres needing synchronised load and stress visibility should accept the modest demand offset for substantially more informative stress predictions than B07 or classical baselines provide.

\subsection{Bangladesh Grid Implications}

Three characteristics of the Bangladesh operational record shape the results. First, nine divisions with co-visible shedding, reserve, and limitation fields in each daily record make composite stress supervision feasible without external telemetry. Second, Dhaka's dominance ($\sim$35.7\% of national demand, highest test-set MAE) means macro-averaged metrics understate capital-region planning sensitivity; error concentration persists across all model families, indicating a data-scale challenge rather than an architecture-specific failure. Third, daily cadence with a seven-day lookback aligns with administrative reporting practice, explaining why shallow recurrent and T-GCN baselines underperform relative to full transformer capacity on the correlation graph: they lack sufficient expressive power to model multi-day seasonal structure within the window while propagating information across densely connected nodes.

\subsection{Practical Deployment Considerations}

A single forecasting service can update regional evening-peak trajectories and a national OSI scalar from one seven-day input window on the same daily cycle as the forecast horizon. Operators could review paired forecasts during day-ahead planning, treating Dhaka as a separately reported diagnostic and monitoring limitation-stack and calendar inputs highlighted by attribution. Shadow-mode operation alongside incumbent tools is advisable before cutover, with division-level MAE thresholds and periodic stress-case review substituting for absent prediction intervals. These recommendations are planning diagnostics grounded in offline accuracy---not demonstrated reductions in grid reliability, economic outcome, or deployment success.

\subsection{Limitations and Future Work}

Evidence is bounded to a single Bangladesh timeline with daily cadence and frozen internal benchmarks; no external replication or field deployment trial is included. Inferential support applies to demand MAE; OSI gains lack hypothesis tests or prediction intervals. Attribution describes model behaviour on the evaluation sample, not physical causation, with documented cross-method discordance ($\rho{=}{-}0.564$ on demand). Uncertainty quantification and S2 residual time-series diagnostics were not executed.

Future work could replicate the chronological protocol on comparable South Asian national systems, evaluate adaptive correlation graphs, extend probabilistic forecasting with quantile or conformal heads, and conduct operator-in-the-loop validation of dual-task explanation dashboards. Such studies would separate architecture-specific gains from country-specific feature geometry without altering the frozen conclusions reported here.
